---
title: "3.1 & 3.2: Phylogency Exercises"
output: html_notebook
---

By Lena Kogan

Necessary packages:

```{r}
install.packages(ape)
install.packages("devtools")
devtools::install_github("ropensci/brranching")
library("ape")
library("brranching")
```

### MINI EXERCISE 1: Understanding Newick and Phylo Objects

```{r}
newick_str = "(((Humans, Chimps), Gorillas), Orangs);"
tr_primates = read.tree(text=newick_str)
plot(tr_primates)
```

Notes about the following commmands:

tr_primates : yields labels for the tips

str("tr_primates") : prints "chr "tr_primates" (the tree name) because of the quotations

str(tr_primates) : this yields information about the tree, like the number of edges, nodes, and tip, as well as class and order

attributes(tr_primates) : this yields details like name, class, and order that are important to the organization of the tree. 

names(tr_primates) : this yields the specific labels that are part of the name detail.

To finish up Mini-Exercise 1, write new code with a new Newick string variable, read the string using read.tree, and plot it. The goal is to generate a four-taxon tree with a new topology. It will be an erroneous tree, depicting relationships that contradict today's biological knowledge. That's okay; this is just an exercise.

```{r}
new_tips = "(Banana, ((Dragonfruit, Pomegranate), Lemon));"
fruits = read.tree(text=new_tips)
plot(fruits)
```

## MINI-EXERCISE 2

1. Copy, paste and re-write a chunk of code (or write it fresh): write a new Newick tree as as string variable, read it as a tree, and plot it. Make sure it is consistent with 2018 botanical knowledge about these four taxa. You may need to modify topology.

```{r}
plant_tips = "(((buckwheat, rhubarb, sorrel), asterids));"
plant_tree = read.tree(text=plant_tips)
plot(plant_tree)
```

2. To these four taxa, add two additional taxa: radish and sequoia. For these six tip taxa, write a new Newick variable, read it, and plot it. Then repeat. Aim to write two alternative trees: one erroneous and one correct. 

Correct tree:
```{r}
correct = "((((buckwheat, rhubarb, sorrel), radish)), (asterids), (Sequoia));"
correct_tree = read.tree(text=correct)
plot(correct_tree)
```

Erroneous tree:
```{r}
erroneous = "((((buckwheat, rhubarb, radish), asterids)), (sorrel), (Sequoia));"
erroneous_tree = read.tree(text=erroneous)
plot(erroneous_tree)
```


a. Explain what's wrong in your erroneous tree and how you've corrected the error in the other tree.

The tree is wrong because it places radish in a monophyletic group with rhubarb and buckwheat, while sorrel is downgraded to hanging out on the outskirts, even though sorrel is more related to buckwheat and rhubarb than radishes are.

b. Use commands such as "str" or "attribute" and comment about whether your erroneous and correct tree differ in tips, Nnodes, or edges. 

```{r}
str(correct_tree)
attributes(correct_tree)
str(erroneous_tree)
attributes(erroneous_tree)
```

Their attributes are the same, but the tip order and number of nodes differ.

## MINI-EXERCISE 3

In November 2017, scientists discovered that there are two Orang species, the known species Pongo abelii and the new species Pongo tapanuliensis. The two taxa are estimated to have separated about 3.5 million years ago. How would you modify the Newick portion of code above to show that there are two species? 

```{r}
newick_str = "(((Humans, Chimps), Gorillas), (Pongo_abelii, Pongo_tapanuliensis));"
tr_primates = read.tree(text=newick_str)
plot(tr_primates)
```

### MINI EXERCISE 4

Practice a few of these data display techniques using the six-taxon phylogeny that includes cabbage, kale, corn and wheat, plus radish and sequoia.

```{r}
plants_two = "(Sequoia, (Wheat,(Corn,(Cabbage, Kale, Radish))));"
plants_two_tree = read.tree(text=plants_two)

plot(plants_two_tree, type="cladogram")
plot(plants_two_tree, type="fan")
plot(plants_two_tree, type="unrooted")
plot(plants_two_tree, type="radial")
```


#### MINI-EXERCISE 5

Do and write up notes about two things:

1. Try generating these trees again, but change from cladogram to phylogram, or to another format for the tree. 

```{r}
# storing data
EG<-c("Abies",
"Ailanthus",
"Aspicllum",
"Atragene",
"Bangia",
"Begonia",
"Berberis",
"Betula",
"Blasia",
"Blechnum",
"Caulerpa",
"Ceratophyllum",
"Chenopodium",
"Clematis",
"Cocculus",
"Corallorrhiza",
"Corchorus",
"Cucurbita",
"Dahlia",
"Dioscorea",
"Dracaena",
"Equisetum",
"Euonymus",
"Euphorbia",
"Fagus",
"Ficus",
"Fontinalis",
"Fossombronia",
"Fritillaria",
"Fucus",
"Gleditschia",
"Gleocapsa",
"Haplomitrium",
"Hedera",
"Helianthus",
"Heterocentron",
"Hippuris",
"Hypoderma",
"Isoetes",
"Linum",
"Liquidambar",
"Marchantia",
"Momordica",
"Monopodium",
"Mucor",
"Nerium",
"Nitella",
"Osmunda",
"Pandanus",
"Papaver",
"Pelargonium",
"Phlomis",
"Phytolacca",
"Pinus",
"Platanus",
"Polytrichum",
"Populus",
"Potamogeton",
"Punctum",
"Quercus",
"Ricinus",
"Rubus",
"Salix",
"Sambucus",
"Sclerotium",
"Selaginella",
"Sempervivum",
"Sponia",
"Stypocaulon",
"Sympodium",
"Taxodium",
"Tecoma",
"Tradescantia",
"Urtica",
"Viola",
"Vitis",
"Wisteria",
"Yucca")

EG_tree <- phylomatic(taxa = EG, get = 'POST')
write.tree(EG_tree, file="eg_tree_stuff")
# cladogram version
plot(EG_tree, no.margin = TRUE, type="cladogram")
# phylogram version
plot(EG_tree, no.margin = TRUE, type="phylogram")
```


2. Then, looking at all your output, what differs most noticably between the pair of trees, and how/why does this relate to the stored tree used to generate them? Try to explain what this means in your own words. 
One of the trees is a cladogram, which has branches of equal length, so while we see ancestry between different taxa, we cannot compare evolutionary time. The second tree plotted (the one with the right angles) corresponds to the command to plot a phylogenetic tree. Like the cladogram, the phylogenetic tree shows ancestry, but also has branches of differing lenghts that correspond to evolutionary time. The phylogenetic tree provides more information, but can be a little trickier to look at.

3. Lastly, look at the output and you'll see that many taxa were excluded. Try looking them up in EOL or Wikipedia. Why was it difficult to include them? What does this say about the history of botany? Are there some things that were excluded that should not have been excluded?  
Some of the taxa, like Sympodium, are not taxa at all, so they cannot be sorted with other plant taxa. Others, like Aspicllum seem to be mispelled. Cleaning the list before organizing it into the cladogram or phylogram would help avoid problems like this. Also, other taxa that are excluded are those that pertain to bryophytes (mosses, non-vascular plants), so they are all in a completely different category, and would not make sense to be included in the tree.

#### FINAL EXERCISE

For the final exercise, I took the list of species from the Welikia database and the list of plants I used in a day from Portfolio 2. Let's read in those csv files and get the species names so we can form our trees:

```{r}
welikia_list = read.csv("lol.csv")
used_list = read.csv("lmao.csv")
welikia_species = paste(welikia_list$Scientific.Name)
used_species = paste(used_list$Scientific.Name)

welikia_species
used_species
```

Now let's format the trees as phylogenies using the phylomatic() method and plot the trees.

```{r}
welikia <- phylomatic(taxa = welikia_species, get = 'POST')
plot(welikia, no.margin = TRUE, type="phylogram")
used <- phylomatic(taxa = used_species, get = 'POST')
plot(used, no.margin = TRUE, type="phylogram")
```

Now let's saved our plotted trees as PDF files so we can view them better!

```{r}
welikia_fn = "welikia_tree.pdf"
pdf(file=welikia_fn)
plot(welikia, no.margin = TRUE, type="phylogram")
dev.off()
cmdstr = paste("open ", welikia_fn, sep="")
system(cmdstr)

used_fn = "plants_i_use_tree.pdf"
pdf(file=used_fn)
plot(used, no.margin = TRUE, type="phylogram")
dev.off()
cmdstr = paste("open ", used_fn, sep="")
system(cmdstr)
```

#### MINI-EXERCISE
******NOT DOING IT BECAUSE I CHOSE 3.4*******
1 What does our list of plants from above look like on a phylogenetic tree? Remember that we assigned as P_resolved_names for the output when we resolved our list of plant genera: "Abies", "Ailanthus", "Begonia", "Berberis", "Betula", "Chenopodium", "Clematis", "Dahlia", "Dracaena", "Equisetum", "Euonymus", "Euphorbia", "Fagus"

+ Modify the code that generated the four-taxon animal phylogeny, putting it into the blank chunk of code below. Then run the code to obtain the output (a phylogeny from Open Tree of Life).


```{r}

```

2. You've developed your own lists of plants to resolve and phylogenize. Use it to Get a phylogeny from OTT.

